---
title: "Analysis for NPR: As Rising Heat Bakes U.S. Cities, The Poor Often Feel It Most"
author: "Roxanne Ready, Theresa Diffendal, Jane Gerard, Jake Gluck and Sean Mussenden | Howard Center for Investigative Journalism"
date: "9/3/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r include=FALSE}
# Save this file and run the following line from the Console to output both HTML and .md formats:
# rmarkdown::render('documentation/NPR-analysis.Rmd', output_format = 'all')
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, paged.print=TRUE)
```

## Introduction
This R markdown document describes the methodology, results and explanation of a portion of the data analysis we conducted in support of reporting project examining the effects of climate-change driven temperature increases on the health of people who live in cities, focusing on Baltimore.  The project was done in partnership with the University of Maryland's Philip Merrill College of Journalism, Capital News Service, the Howard Center for Investigative Journalism, NPR, Wide Angle Youth Media and WMAR.  

* [NPR: As Rising Heat Bakes U.S. Cities, The Poor Often Feel It Most](https://www.npr.org/templates/story/story.php?storyId=754044732&live=1)


## Setup
Before running this file, please view and run the [Code Red Data Cleaning document](https://github.com/smussenden/2019-baltimore-climate-health-project-data-repo/blob/master/documentation/code-red-data-cleaning.md)** for this project. As well as outputting necessary cleaned data for the following ananlysis, that document also includes the following items necessary to understand this analysis: 

* definitions
* source data citation and information
* cleaning methodology 
* software tools used

### Load packages
```{r}
#######################
#### Load Packages ####
#######################
# For debugging rm(list=ls())

library(tidyverse) # For general data science goodness
library(DescTools) # For %like% operator
library(corrr) # For correlation matrices
library(colorspace) # For improved color palettes
library(ggplot2) # For graphing
library(ggrepel) # For graph labeling
library(lubridate) # For working with and mutating dates
library(broom) # for converting lm output to dataframe
require(scales) # For percent labeling on distribution tables
library(TTR) # For moving averages

# Turn off scientific notation in RStudio (prevents coersion to character type)
options(scipen = 999)

```

### Load variables and data
```{r}
#########################
#### Store Variables ####
#########################

#### Common path to output data folder ####
path_to_data <- "../data/output-data/"

###################
#### Load Data ####
###################

### Indoor temperature and humidity sensor data

# Michael and Alberta
folder <- "temperature_sensors/michael/"
michael_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "michael_day_hourly_averages.csv"))
michael_day_minute_averages <- read_csv(paste0(path_to_data, folder, "michael_day_minute_averages.csv"))

# Stephanie and family
folder <- "temperature_sensors/stephanie/"
stephanie_day_minute_averages <- read_csv(paste0(path_to_data, folder, "stephanie_day_minute_averages.csv"))
stephanie_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "stephanie_day_hourly_averages.csv"))

# Audrey
folder <- "temperature_sensors/audrey/"
audrey_day_minute_averages <- read_csv(paste0(path_to_data, folder, "audrey_day_minute_averages.csv"))
audrey_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "audrey_day_hourly_averages.csv"))

# Tammy
folder <- "temperature_sensors/tammy/"
tammy_day_minute_averages <- read_csv(paste0(path_to_data, folder, "tammy_day_minute_averages.csv"))
tammy_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "tammy_day_hourly_averages.csv"))

### Outdoor temperature data 

# Inner Harbor temperature data
folder <- "baltimore_weather_stations/"
dmh <- read_csv(paste0(path_to_data, folder, "dmh.csv"))

# Historical temperature data, Baltimore and U.S. 
folder <- "1895-2019-average-temperature/"
us_change_temp <- read_csv(paste0(path_to_data, folder, "us_change_temp.csv"))
baltimore_change_temp <- read_csv(paste0(path_to_data, folder, "baltimore_change_temp.csv"))

# Heat index projections
folder <- "heat_index_projections/"
heat_index_projections <- read_csv(paste0(path_to_data, folder, "heat_index_projections.csv"))

### Urban heat island, tree canopy, demographics data
folder <- "tree_temp_demographics/"

# Neighborhood geography
nsa_tree_temp <- read_csv(paste0(path_to_data, folder, "nsa_tree_temp.csv"))

# Community statistical area geography
csa_tree_temp_demographics <- read_csv(paste0(path_to_data, folder, "csa_tree_temp_demographics.csv"))

# Blocks geography
blocks_tree_temp_demographics <- read_csv(paste0(path_to_data, folder, "blocks_tree_temp_demographics.csv"))


### Hospital Data
folder <- "hospital_data/"

# Inpatient admissions data
ip_full_zip_correlation_matrix <- read_csv(paste0(path_to_data, folder, "ip/ip_full_zip_correlation_matrix.csv"))
ip_full_zip_medicaid_correlation_matrix <- read_csv(paste0(path_to_data, folder, "ip/ip_full_zip_medicaid_correlation_matrix.csv"))

# Emergency room admissions data
op_er_full_zip_correlation_matrix <- read_csv(paste0(path_to_data, folder, "op_er/op_er_full_zip_correlation_matrix.csv"))
op_er_full_zip_medicaid_correlation_matrix <- read_csv(paste0(path_to_data, folder, "op_er/op_er_full_zip_medicaid_correlation_matrix.csv"))

### EMS Data
folder <- "ems/"
dmh_ems <- read_csv(paste0(path_to_data, folder, "dmh_ems.csv")) 
EMS_all <- read_csv(paste0(path_to_data, folder, "EMS_all.csv")) 

```

## Line-by-Line Fact Check

### Fact: Franklin Square heat [cq - audio and web]
**Audio Story:** "Her neighborhood -- Franklin Square, no relation to her name -- is hotter than about two thirds of the neighborhoods in Baltimore."

**Web Story:** "Her neighborhood, Franklin Square, is hotter than about two thirds of the neighborhoods in Baltimore – about six degrees hotter than the city's coolest neighborhood."

#### Explanation [cq]
Using the urban heat island measurements taken in August 2018 showing block-by-block temperature variations, we calculated median afternoon temperatures for each of the city's 278 neighborhoods. At 95.4 degrees, Franklin Square was warmer than 67 percent (about two-thirds) of Baltimore's neighborhoods. The city's coolest neighborhood -- "Gwynns Falls/Leakin Park" -- was 89.3 degrees F, 6.1 degrees less than Franklin Square.

#### Supporting code and output [cq]
```{r}
# Rank, neighborhoods cooler 
nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  mutate(rank = dense_rank(temp_median_aft), pct_nhoods_cooler = (rank/max(rank))*100) %>%
  filter(nsa_name == "franklin square")

# Difference between Franklin Square and coolest
nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter((nsa_name == "franklin square") | (temp_median_aft == min(temp_median_aft))) %>%
  spread(nsa_name, temp_median_aft) %>%
  mutate(difference = `franklin square` - `gwynns falls/leakin park`)



```

### Fact: Franklin Square poverty [cq - web and audio]
**Audio Story:** "It's also in one of the city's poorer areas."

**Web Story:** "It's also in one of the city's poorest communities, with more than one third of residents living in poverty."

#### Explanation [cq]
Franklin Square is entirely contained by a "community statistical area" known as "Southwest Baltimore". The poverty rate here is 36 percent, or more than one-third. There are 55 CSAs in Baltimore.  Southwest Baltimore has the 51st highest poverty rate, making it poorer than more than 90 percent of CSAs.

#### Supporting code and output [cq]
```{r}
csa_tree_temp_demographics %>%
  select(`csa2010`, percent_of_family_households_living_below_the_poverty_line) %>%
  mutate(rank = dense_rank(percent_of_family_households_living_below_the_poverty_line), pct_nhoods_wealthier = (rank/max(rank))) %>%
  filter(`csa2010` == "southwest baltimore")
```
### Fact: Heat and poverty [cq - web and audio]
**Audio Story:** "And the hottest parts of the city also have higher rates of poverty."

**Web Story:** "Across Baltimore, the hottest areas tend to be the poorest and that pattern is not unusual."

#### Explanation [cq]
In Baltimore's "community statistical areas", we examined the relationship between heat (median afternoon temperature in our urban heat island data) and poverty.  Where an r of 1 would indicate a perfect positive linear relationship and an r of -1 would indicate a perfect negative linear relationship and 0 indicating no relationship. There were moderate correlations with poverty (r =.38).

#### Supporting code and output [cq]
```{r}
csa_tree_temp_demographics %>%
  select_if(is.numeric) %>%
  as.matrix() %>%
  correlate() %>%
  focus(matches("temp_")) %>%
  mutate(variable=rowname) %>%
  select(variable, temp_median_aft) %>%
  filter(variable=="percent_of_family_households_living_below_the_poverty_line")

```

### Fact: Neighborhood heat difference [cq - web and audio]
**Audio Story:** "Citywide in Baltimore, the hottest neighborhoods can differ by as much as 10 degrees from the coolest."  

**Web Story:** "According to a Howard Center analysis of U.S. Census data and air temperature data obtained from Portland State University and the Science Museum of Virginia, the hottest neighborhoods in Baltimore can differ by as much as 10 degrees from the coolest."

#### Explanation 
Using median afternoon temperature in a 2018 study by researchers analyzing the city's urban heat island, there was a 9.98 degree difference between the city's hottest neighborhood (McElderry Park) and coolest neighborhood (Gwynns Falls/Leakin Park). 

#### Supporting code and output 
```{r}
nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter((temp_median_aft == min(temp_median_aft)) | (temp_median_aft == max(temp_median_aft)))

nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter((temp_median_aft == min(temp_median_aft)) | (temp_median_aft == max(temp_median_aft))) %>%
  spread(nsa_name, temp_median_aft) %>%
  mutate(difference = `mcelderry park` - `gwynns falls/leakin park`)

```

### Fact: EMS calls for heat stroke [cq - web and audio]
**Audio Story:** "When the heat index reached dangerous levels last summer, EMS calls increased citywide for heat stroke."

**Web Story:** "In the summer of 2018 in Baltimore, when the heat index reached 103 degrees — the threshold deemed dangerous by the National Weather Service — EMS calls increased dramatically citywide for potentially fatal heat stroke."

#### Explanation [cq]
Using emergency medical call records from Baltimore City, we examined calls during Summer 2018.  They were aligned to heat index data captured at the Inner Harbor and adjusted for the urban heat island using the ZIP Code of each call location.  The statistics in the table below represent the number of hours that passed between calls for select conditions when the temperature was in a given heat index bucket. 

As expected, calls for heat stroke increased dramatically.  When it was under 80 degrees, there was a call for heat stroke every 480 hours, or about once every thre weeks. When it was 103 degrees or higher, there was a call for heat stroke every 1.4 hours.

#### Supporting Code [cq]
```{r}

# Select conditions
conditions <- c("Heat Exhaustion/Heat Stroke")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  tidyr::spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`)

```

### Fact: EMS calls for chronic conditions [Audio cq, First web graf cq, Second web graf not cq]
**Audio Story:** "But calls also increased for chronic conditions, including several cardiovascular and respiratory conditions." 

**Web Story:** "But calls increased for chronic conditions too: EMS calls for chronic obstructive pulmonary disorder (COPD) increased by nearly 70 percent. Calls for respiratory distress increased by 20 percent. Calls for cardiac arrest rose by 80 percent and those for high blood pressure more than doubled. Other conditions also spiked: Psychiatric disorders, substance abuse and dehydration, among others."

**Web Story:** " And living day after day in an environment that's literally hotter isn't just uncomfortable, it can have dire and sometimes deadly health consequences – a fact we found reflected in Baltimore's soaring rates of emergency calls when the heat index spiked to dangerous levels." 

#### Explanation [cq]
By aligning emergency medical calls in Baltimore in summer 2018 for select conditions that are affected by heat with the heat index in Baltimore at the time of call, we were able to compare how rates differed when the heat index was below 80 degrees and above 103, the level the NWS defines as "dangerous". The second column reflects the increase in calls when the temperature rose.

#### Supporting code and output [cq]
```{r}

# Select conditions
conditions <- c("Dehydration","Respiratory Distress", "COPD (Emphysema/Chronic Bronchitis)", "Cardiac Arrest", "CHF (Congestive Heart Failure)", "Behavioral/Psychiatric Disorder", "Dehydration", "Chest Pain - STEMI","Hypertension","Substance/Drug Abuse","Withdrawal/Overdose Drugs","Withdrawal/Overdose ETOH")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  tidyr::spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`) %>%
  mutate(`calls_per_day_under_80` = 24/`not_unsafe_under_80`) %>%
  mutate(`calls_per_day_over_103` = 24/`danger_103_124`) %>%
  mutate(difference_day_percent = ((`calls_per_day_over_103`-`calls_per_day_under_80`)/`calls_per_day_under_80`)) %>%
  select(primary_impression_group, difference_day_percent, everything())


```

### Fact: Medicaid and chronic conditions [cq - web and audio]
**Audio Story:** "And even when controlling for income, there were differences across the city. From 2013 to 2018, Medicaid patients in Baltimore's hottest areas visited the hospital with those conditions at higher rates than Medicaid patients in the cooler areas, according to the Howard Center analysis."

**Web Story:** "The heat affected residents citywide, but even when controlling for income by only looking at the patterns of Medicaid patients, there were differences across the city. From 2013 to 2018, Medicaid patients in Baltimore's hottest areas visited the hospital at higher rates than Medicaid patients in the city's coolest areas. The low-income patients in the city's hot spots visited more often with several conditions, including asthma, COPD and heart disease, according to hospital inpatient and emergency room admissions data from the state's Health Services Cost Review Commission."

#### Explanation [cq]

We examined rates of chronic conditions among low-income people in different parts of Baltimore by examining in-patient hospital admissions by people on Medicaid in Baltimore, and discovered that low-income people in different parts of the city had different prevalance rates for chronic conditions affected by heat -- asthma, COPD, heart disease, kidney disease and diabetes. And, we found, those differences varied in line with temperature differences in the area in which they lived.

There were moderate to strong positive relationships (copd, r=.75; asthma, r=.52; heart_disease, r=.71; diabetes, r=.5) between a ZIP code's prevalance rate for chronic medical conditions as diagnosed in inpatient hospital visits and a ZIP code's median afternoon temperature as measured by urban heat island researchers in August 2018. There were moderate to strong positive relationships (copd, r=.51; asthma, r=.38; heart_disease, r=.44; diabetes, r=.44) between a ZIP code's prevalance rate for chronic medical conditions as diagnosed in emergency room visits and a ZIP code's median afternoon temperature as measured by urban heat island researchers in August 2018. That is to say: the higher the neighborhood temperature, the higher the disease rate amongst the poorest inhabitants, and vice versa. This is not a causal relationship we are describing.  

#### Supporting Code [cq]
```{r}
ip_full_zip_medicaid_correlation_matrix %>%
    filter(str_detect(rowname, "asthma|copd|heart_disease|diabetes")) %>%
    select(rowname, temp_median_aft)

op_er_full_zip_medicaid_correlation_matrix %>%
    filter(str_detect(rowname, "asthma|copd|heart_disease|diabetes")) %>%
    select(rowname, temp_median_aft)

```

### Fact: Franklin Square tree canopy [cq - web and audio]
**Audio Story:** "The neighborhood where Shakira Franklin lives has increased its tree canopy over time. But in recent years, it's still been among the city's lowest.

**Web Story:** "The neighborhood where Shakira Franklin lives has increased its tree canopy over time. But by 2015, it still was among the city's lowest."

#### Explanation [cq]
In 2007, 14.2 percent of the neighborhood was covered by tree canopy in the summer.  By 2015, that had increased to 16.1 percent, a 1.9 percentage point increase. In 2015, two-thirds of Baltimore's 278 neighborhoods had more tree canopy than Franklin Square's 16 percent coverage. 

#### Supporting Code [cq]
```{r}

# change
nsa_tree_temp %>%
  select(nsa_name, `avg_canopy_%_07` = `07_lid_mean`, `avg_canopy_15` = `15_lid_mean`, `%_point_change` = lid_change_percent_point) %>%
  filter(nsa_name == "franklin square")

# rank
nsa_tree_temp %>%
  select(nsa_name, `avg_canopy_15` = `15_lid_mean`) %>%
  arrange(nsa_name) %>%
  mutate(rank = dense_rank(desc(`avg_canopy_15`)), pct_nhoods_w_more_tree_canopy = (rank/max(rank))) %>%
  filter(nsa_name == "franklin square")


```

## Mental illness factcheck

### Fact: Rate of increase of psychiatric calls [cq] 

"The Howard Center analyzed data from emergency response calls in Baltimore. They found that in the summer of 2018, calls for psychiatric conditions increased by nearly 40 percent when the heat index spiked above 103."

#### Explanation [cq]
Using emergency medical call records from Baltimore City, we examined calls during Summer 2018.  They were aligned to heat index data captured at the Inner Harbor and adjusted for the urban heat island using the ZIP Code of each call location.  

In Summer 2018, when the heat index was under 80 degrees, there was a medical call for a behavioral and psychiatric disorder every 1.77 hours (1 hour, 46 minutes). When the heat index hit 103 degrees, the rate of calls increased dramatically -- to one call every 1.29 hours (1 hour, 17 minutes). 

That's an increase in the rate of 29 minutes, or an increase of 37 percent. Yes, it's weird to say that a *decrease* in the number of hours between calls actually represents an increase in the rate, but it makes sense logically.  If there are fewer hours between calls, then there are more calls on any given day.  It's a 37 percent change. 

It's easier to see this if we convert the hours between calls statistic to calls per day by dividing the hours between calls by 24.  We get a rate of calls per day of 13.55 when it's under 80, and a rate of 18.56 when it's over 103.  The percent change between 13.55 and 18.56 is 37 percent. 

#### Supporting code and output [cq]
``` {r}

# Select conditions
conditions <- c("Behavioral/Psychiatric Disorder")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`) %>%
  mutate(percent_change = (`not_unsafe_under_80`-`danger_103_124`)/`danger_103_124`) %>%
  mutate(`calls_per_day_under_80` = 24/`not_unsafe_under_80`) %>%
  mutate(`calls_per_day_over_103` = 24/`danger_103_124`) %>%
  mutate(difference_day_percent = ((`calls_per_day_over_103`-`calls_per_day_under_80`)/`calls_per_day_under_80`))


```


### Fact: Substance abuse calls rate [cq]
"Calls relating to substance abuse more than doubled in extreme heat, according to data analyzed by the Howard Center."

#### Explanation [cq]
Using emergency medical call records from Baltimore City, we examined calls during Summer 2018.  They were aligned to heat index data captured at the Inner Harbor and adjusted for the urban heat island using the ZIP Code of each call location.  

In Summer 2018, when the heat index was under 80 degrees, there was a medical call for substance abuse every 2.96 hours (nearly 3 hours). When the heat index hit 103 degrees, the rate of calls increased dramatically -- to one call every 1.35 hours. 

That's an increase in the rate of about an hour and a half, or an increase of 118 percent. Yes, it's weird to say that a *decrease* in the number of hours between calls actually represents an increase in the rate, but it makes sense logically.  If there are fewer hours between calls, then there are more calls on any given day.  We use the classic percent change formula ((new-old)/old) to calculate the difference between the two values. It's a 118 percent change. 

It's easier to see this if we convert the hours between calls statistic to calls per day by dividing the hours between calls by 24.  We get a rate of calls per day of 8.11 when it's under 80, and a rate of 17.67 when it's over 103.  The percent change between those values is 118 percent.

#### Supporting code and output [cq]
``` {r}

# Select conditions
conditions <- c("Substance/Drug Abuse")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`) %>%
  mutate(percent_change = (`not_unsafe_under_80`-`danger_103_124`)/`danger_103_124`) %>%
  mutate(`calls_per_day_under_80` = 24/`not_unsafe_under_80`) %>%
  mutate(`calls_per_day_over_103` = 24/`danger_103_124`) %>%
  mutate(difference_day_percent = ((`calls_per_day_over_103`-`calls_per_day_under_80`)/`calls_per_day_under_80`))


```

-30-