---
title: "Analysis for NPR story"
author: "Roxanne Ready, Theresa Diffendal, Jane Gerard, Jake Gluck and Sean Mussenden | Howard Center for Investigative Journalism"
date: "9/3/2019"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    df_print: paged
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r include=FALSE}
# Save this file and run the following line from the Console to output both HTML and .md formats:
# rmarkdown::render('documentation/neighborhoods-data-analysis.Rmd', output_format = 'all')
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, paged.print=TRUE)
```

## Introduction

This R markdown document describes the methodology, results and explanation of a portion of the data analysis we conducted in support of reporting project examining the effects of climate-change driven temperature increases on the health of people who live in cities, focusing on Baltimore.  The project was done in partnership with the University of Maryland's Philip Merrill College of Journalism, Capital News Service, the Howard Center for Investigative Journalism, NPR, Wide Angle Youth Media and WMAR.  
XXXX PUT IN LINKS TO PLACES HERE

Read the whole project, "[Code Red: Baltimore's Climate Divide](https://cnsmaryland.org/code-red
)." 

This document provides support for facts based on data analysis in the story, "[In urban heat islands, climate crisis hits harder](https://cnsmaryland.org/interactives/summer-2019/code-red/neighborhood-heat-inequality.html#story)" as well as LINK TO NPR STORY. 

## Setup
XXX UPDATE LINK
Before running this file, please view and run the [Code Red Data Cleaning document](https://github.com/smussenden/2019-baltimore-climate-health-project-data-repo/blob/master/documentation/code-red-data-cleaning.md)** for this project. As well as outputting necessary cleaned data for the following ananlysis, that document also includes the following items necessary to understand this analysis: 

* definitions
* source data citation and information
* cleaning methodology 
* software tools used

### Load packages

```{r}
#######################
#### Load Packages ####
#######################
# For debugging rm(list=ls())

library(tidyverse) # For general data science goodness
library(DescTools) # For %like% operator
library(corrr) # For correlation matrices
library(colorspace) # For improved color palettes
library(ggplot2) # For graphing
library(ggrepel) # For graph labeling
library(lubridate) # For working with and mutating dates
library(broom) # for converting lm output to dataframe
require(scales) # For percent labeling on distribution tables
library(TTR) # For moving averages

# Turn off scientific notation in RStudio (prevents coersion to character type)
options(scipen = 999)

```

### Load variables and data


```{r}
#########################
#### Store Variables ####
#########################

#### Common path to output data folder ####
path_to_data <- "../data/output-data/"

###################
#### Load Data ####
###################

### Indoor temperature and humidity sensor data

# Michael and Alberta
folder <- "temperature_sensors/michael/"
michael_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "michael_day_hourly_averages.csv"))
michael_day_minute_averages <- read_csv(paste0(path_to_data, folder, "michael_day_minute_averages.csv"))

# Stephanie and family
folder <- "temperature_sensors/stephanie/"
stephanie_day_minute_averages <- read_csv(paste0(path_to_data, folder, "stephanie_day_minute_averages.csv"))
stephanie_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "stephanie_day_hourly_averages.csv"))

# Audrey
folder <- "temperature_sensors/audrey/"
audrey_day_minute_averages <- read_csv(paste0(path_to_data, folder, "audrey_day_minute_averages.csv"))
audrey_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "audrey_day_hourly_averages.csv"))

# Tammy
folder <- "temperature_sensors/tammy/"
tammy_day_minute_averages <- read_csv(paste0(path_to_data, folder, "tammy_day_minute_averages.csv"))
tammy_day_hourly_averages <- read_csv(paste0(path_to_data, folder, "tammy_day_hourly_averages.csv"))

### Outdoor temperature data 

# Inner Harbor temperature data
folder <- "baltimore_weather_stations/"
dmh <- read_csv(paste0(path_to_data, folder, "dmh.csv"))

# Historical temperature data, Baltimore and U.S. 
folder <- "1895-2019-average-temperature/"
us_change_temp <- read_csv(paste0(path_to_data, folder, "us_change_temp.csv"))
baltimore_change_temp <- read_csv(paste0(path_to_data, folder, "baltimore_change_temp.csv"))

# Heat index projections
folder <- "heat_index_projections/"
heat_index_projections <- read_csv(paste0(path_to_data, folder, "heat_index_projections.csv"))

### Urban heat island, tree canopy, demographics data
folder <- "tree_temp_demographics/"

# Neighborhood geography
nsa_tree_temp <- read_csv(paste0(path_to_data, folder, "nsa_tree_temp.csv"))

# Community statistical area geography
csa_tree_temp_demographics <- read_csv(paste0(path_to_data, folder, "csa_tree_temp_demographics.csv"))

# Blocks geography
blocks_tree_temp_demographics <- read_csv(paste0(path_to_data, folder, "blocks_tree_temp_demographics.csv"))


### Hospital Data
folder <- "hospital_data/"

# Inpatient admissions data
ip_full_zip_correlation_matrix <- read_csv(paste0(path_to_data, folder, "ip/ip_full_zip_correlation_matrix.csv"))
ip_full_zip_medicaid_correlation_matrix <- read_csv(paste0(path_to_data, folder, "ip/ip_full_zip_medicaid_correlation_matrix.csv"))

# Emergency room admissions data
op_er_full_zip_correlation_matrix <- read_csv(paste0(path_to_data, folder, "op_er/op_er_full_zip_correlation_matrix.csv"))
op_er_full_zip_medicaid_correlation_matrix <- read_csv(paste0(path_to_data, folder, "op_er/op_er_full_zip_medicaid_correlation_matrix.csv"))

### EMS Data
folder <- "ems/"
dmh_ems <- read_csv(paste0(path_to_data, folder, "dmh_ems.csv")) 
EMS_all <- read_csv(paste0(path_to_data, folder, "EMS_all.csv")) 

####################################
######## Define Functions ##########
####################################

# Function to save each matrix as CSV
write_matrix_csv <- function(dataframe) {
  # Store dataframe name for later use
  dataframe_name <- deparse(substitute(dataframe))
  
  # Create filename for csv
  filename <- paste0("data/output-data/correlation_matrices/", dataframe_name,".csv")
  
  # Write out csv  
  write_csv(dataframe, path = filename)
  
} 

# Function to make a nice little correlation matrix heatmap for each graphic

make_correlation_matrix_graphic <- function(dataframe, grouping = "GROUPING") {
  
  # Store name of dataframe for use in title
  dataframe_name <- deparse(substitute(dataframe))
  
  # Build chart title
  chart_title <- paste0("Correlations by ", grouping, " in Baltimore City | ", dataframe_name )
  
  # Create graph
  ggplot(data = dataframe, aes(x = variable_2, y = variable)) +
    geom_tile(aes(fill = value)) +
    scale_fill_gradient2(low = "blue", high = "red", mid="white", midpoint=0) +
    geom_text(aes(label = round(value, 2)*100), size = 10) +
    ggtitle(chart_title) +
    theme(axis.text=element_text(size=14),
          axis.text.x = element_text(size=14,angle=50,hjust=1),
          plot.title = element_text(size=14)
    )
  # Create filename and filepath to save image. 
  filename <- paste0(dataframe_name,".png")
  ggsave(filename, plot = last_plot(), device = "png", path = "data/output-data/plots/correlation-matrix-images", scale = 1, width = 20, height = 20, units = "in", dpi = 300)
  
}  

select_x <- function(df){
  return(df %>%
           select_if(is.numeric) 
           # select(-matches("objectid"), 
           #        -matches("csa2010"), 
           #        -matches("id"), 
           #        -matches("09"), 
           #        -matches("1718"), 
           #        -matches("change_percent")
           #        )
         )
}

```

### Fact: Franklin Square heat [cq]
Audio story: "Her neighborhood -- Franklin Square, no relation to her name -- is hotter than about two thirds of the neighborhoods in Baltimore."
Web story: "Her neighborhood, Franklin Square, is hotter than about two thirds of the neighborhoods in Baltimore – about five degrees hotter than the city's coolest neighborhood."

#### Explanation [cq]
Using the urban heat island measurements taken in August 2018 showing block-by-block temperature variations, we calculated median afternoon temperatures for each of the city's 278 neighborhoods. At 95.4 degrees, Franklin Square was warmer than 67 percent (two-thirds) of Baltimore's neighborhoods. The city's coolest neighborhood -- excluding Leakin Park, which isn't really a neighborhood, per se -- was Dickeyville, at 90.3 degrees F, 5.1 degrees less than Franklin Square.

#### Supporting code and output [cq]
```{r}
# Rank 
nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter(nsa_name != "gwynns falls/leakin park") %>%
  mutate(rank = dense_rank(temp_median_aft), pct_nhoods_cooler = (rank/max(rank))) %>%
  filter(nsa_name == "franklin square")

nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter(nsa_name != "gwynns falls/leakin park") %>%
  filter((nsa_name == "franklin square") | (temp_median_aft == min(temp_median_aft)))



```

### Fact: Franklin Square poverty [cq]
Audio story: "It's also in one of the city's poorer areas."
Web story: "It's also in one of the city's poorest communities, with more than one third of residents living in poverty."

#### Explanation [cq]
Franklin Square is entirely contained by a "community statistical area" known as "Southwest Baltimore". The poverty rate here is 36 percent, or more than one-third. There are 55 CSAs in Baltimore.  Southwest Baltimore has the 51st highest poverty rate, making it poorer than more than 90 percent of CSAs.

#### Supporting code and output [cq]
```{r}
csa_tree_temp_demographics %>%
  select(`csa2010`, percent_of_family_households_living_below_the_poverty_line) %>%
  mutate(rank = dense_rank(percent_of_family_households_living_below_the_poverty_line), pct_nhoods_wealthier = (rank/max(rank))) %>%
  filter(`csa2010` == "southwest baltimore")
```
# Fact: Heat and poverty [cq]
Audio story: "And the hottest parts of the city also have higher rates of poverty."
Web story: "Across Baltimore, the hottest areas tend to be the poorest and that pattern is not unusual."

#### Explanation [cq]
In Baltimore's "community statistical areas", we examined the relationship between heat (median afternoon temperature in our urban heat island data) and poverty.  Where an r of 1 would indicate a perfect positive linear relationship and an r of -1 would indicate a perfect negative linear relationship and 0 indicating no relationship. There were moderate correlations with poverty (r =.38).

#### Supporting code and output [cq]
```{r}
csa_tree_temp_demographics %>%
  select_x() %>%
  as.matrix() %>%
  correlate() %>%
  focus(matches("temp_")) %>%
  mutate(variable=rowname) %>%
  select(variable, temp_median_aft) %>%
  filter(variable=="percent_of_family_households_living_below_the_poverty_line")

```

### Fact: Neighborhood heat difference///NEEDS WORK
Audio story: "Citywide in Baltimore, the hottest neighborhoods can differ by as much as 10 degrees from the coolest."
Web story: "According to a Howard Center analysis of U.S. Census data and air temperature data obtained from Portland State University and the Science Museum of Virginia, some of the poorest parts of Baltimore could get as much as 9 degrees hotter than the coolest areas." ///NEEDS WORK

#### Explanation 
Using median afternoon temperature in a 2018 study by researchers analyzing the city's urban heat island, there was a 9 degree difference between the city's hottest neighborhood and coolest neighborhood. If we leave in Gwynns Falls/Leakin Park, the difference is 10 degrees.  


```{r}
nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter(nsa_name != "gwynns falls/leakin park") %>%
  filter((temp_median_aft == min(temp_median_aft)) | (temp_median_aft == max(temp_median_aft)))

nsa_tree_temp %>%
  select(nsa_name, temp_median_aft) %>%
  filter((temp_median_aft == min(temp_median_aft)) | (temp_median_aft == max(temp_median_aft)))

```

### Fact: EMS calls for heat stroke [cq]
Audio story: "When the heat index reached dangerous levels last summer, EMS calls increased citywide for heat stroke."
Web story: "In the summer of 2018 in Baltimore, when the heat index reached 103 degrees — the threshold deemed dangerous by the National Weather Service — EMS calls increased dramatically citywide for potentially fatal heat stroke."

#### Explanation [cq]

Using emergency medical call records from Baltimore City, we examined calls during Summer 2018.  They were aligned to heat index data captured at the Inner Harbor and adjusted for the urban heat island using the ZIP Code of each call location.  The statistics in the table below represent the number of hours that passed between calls for select conditions when the temperature was in a given heat index bucket. 

As expected, calls for heat stroke increased dramatically.  When it was under 80 degrees, there was a call for heat stroke every 480 hours, or about once every thre weeks. When it was 103 degrees or higher, there was a call for heat stroke every 1.4 hours.

#### Supporting Code
```{r}

# Select conditions
conditions <- c("Heat Exhaustion/Heat Stroke")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  tidyr::spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`)

```

### Fact: EMS calls for chronic conditions 
Audio story: "But calls also increased for chronic conditions, including several cardiovascular and respiratory conditions."
Web story: "But calls increased for chronic conditions too: EMS calls for chronic obstructive pulmonary disorder (COPD) increased by nearly 70 percent. Calls for respiratory distress increased by 20 percent. Calls for cardiac arrest rose by 80 percent and those for high blood pressure more than doubled. Other conditions also spiked: Psychiatric disorders, substance abuse and dehydration, among others."
Web story: "And living day after day in an environment that's literally hotter isn't just uncomfortable, it can have dire and sometimes deadly health consequences – a fact we found reflected in Baltimore's soaring rates of emergency calls and hospital admissions when the heat index spiked to dangerous levels."///ISSUE WITH HOSPTIAL ADMISSIONS

#### Explanation [cq]
By aligning emergency medical calls in Baltimore in summer 2018 for select conditions that are affected by heat with the heat index in Baltimore at the time of call, we were able to compare how rates differed when the heat index was below 80 degrees and above 103, the level the NWS defines as "dangerous". The second column reflects the increase in calls when the temperature rose.

#### Supporting code and output [cq]
```{r}

# Select conditions
conditions <- c("Dehydration","Respiratory Distress", "COPD (Emphysema/Chronic Bronchitis)", "Cardiac Arrest", "CHF (Congestive Heart Failure)", "Behavioral/Psychiatric Disorder", "Dehydration", "Chest Pain - STEMI","Hypertension","Substance/Drug Abuse","Withdrawal/Overdose Drugs","Withdrawal/Overdose ETOH")

# Calculate the total number of hours over the course of Summer 2018 that the heat index fell into each heat index level, as defined by the national weather service: not unsafe (under 80), caution (80-89), extreme caution (90-102), danger (103-124).   

heat_index_count_per_nws_five_scale_bucket <- dmh_ems %>%
  select(heat_index_nws_five_scale_bucket) %>%
  group_by(heat_index_nws_five_scale_bucket) %>%
  summarise(heat_index_count_per_nws_five_scale_bucket=n()) %>%
  arrange(heat_index_nws_five_scale_bucket)

# For each target condition, calculate the number of hours between calls at each temperature level.  This metric allows us to account for the fact that simply counting calls in each bucket would be flawed, because it wouldn't adjust for the rarity of very hot temperatures. 

EMS_all %>%
  filter(primary_impression_group %in% conditions) %>%
  group_by(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket) %>%
  summarise(condition_calls_count_per_bucket=n()) %>%
  inner_join(heat_index_count_per_nws_five_scale_bucket, by = c("adjusted_heat_index_nws_five_scale_bucket" = "heat_index_nws_five_scale_bucket")) %>%
  mutate(hours_per_call = heat_index_count_per_nws_five_scale_bucket/condition_calls_count_per_bucket) %>%
  select(primary_impression_group, adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  tidyr::spread(adjusted_heat_index_nws_five_scale_bucket, hours_per_call) %>%
  select(primary_impression_group, `not_unsafe_under_80`,`danger_103_124`) %>%
  mutate(`calls_per_day_under_80` = 24/`not_unsafe_under_80`) %>%
  mutate(`calls_per_day_over_103` = 24/`danger_103_124`) %>%
  mutate(difference_day_percent = ((`calls_per_day_over_103`-`calls_per_day_under_80`)/`calls_per_day_under_80`)) %>%
  select(primary_impression_group, difference_day_percent, everything())


```

### Fact: Medicaid and chronic conditions
Audio story: "And even when controlling for income, there were differences across the city. From 2013 to 2018, Medicaid patients in Baltimore's hottest areas visited the hospital with those conditions at higher rates than Medicaid patients in the cooler areas, according to the Howard Center analysis."
Web story: "The heat affected residents citywide, but even when controlling for income by only looking at the patterns of Medicaid patients, there were differences across the city. From 2013 to 2018, Medicaid patients in Baltimore's hottest areas visited the hospital at higher rates than Medicaid patients in the city's coolest areas. The low-income patients in the city's hot spots visited more often with several conditions, including asthma, COPD and heart disease, according to hospital inpatient and emergency room admissions data from the state's Health Services Cost Review Commission."

#### Explanation

We examined rates of chronic conditions among low-income people in different parts of Baltimore by examining in-patient hospital admissions by people on Medicaid in Baltimore, and discovered that low-income people in different parts of the city had different prevalance rates for chronic conditions affected by heat -- asthma, COPD, heart disease, kidney disease and diabetes. And, we found, those differences varied in line with temperature differences in the area in which they lived.

There were moderate to strong positive relationships (copd, r=.75; asthma, r=.52; heart_disease, r=.71; diabetes, r=.5) between a ZIP code's prevalance rate for chronic medical conditions as diagnosed in inpatient hospital visits and a ZIP code's median afternoon temperature as measured by urban heat island researchers in August 2018. There were moderate to strong positive relationships (copd, r=.51; asthma, r=.38; heart_disease, r=.44; diabetes, r=.44) between a ZIP code's prevalance rate for chronic medical conditions as diagnosed in emergency room visits and a ZIP code's median afternoon temperature as measured by urban heat island researchers in August 2018. That is to say: the higher the neighborhood temperature, the higher the disease rate amongst the poorest inhabitants, and vice versa. This is not a causal relationship we are describing.  

#### Supporting Code

```{r}
ip_full_zip_medicaid_correlation_matrix %>%
    filter(str_detect(rowname, "asthma|copd|heart_disease|diabetes")) %>%
    select(rowname, temp_median_aft)

op_er_full_zip_medicaid_correlation_matrix %>%
    filter(str_detect(rowname, "asthma|copd|heart_disease|diabetes")) %>%
    select(rowname, temp_median_aft)

```

### Fact: Franklin Square tree canopy
Audio story: "The neighborhood where Shakira Franklin lives has increased its tree canopy over time. But in recent years, it's still been among the city's lowest.
Web story: "The neighborhood where Shakira Franklin lives has increased its tree canopy over time. But by 2015, it still was among the city's lowest."

#### Explanation
In 2007, 14.2 percent of the neighborhood was covered by tree canopy in the summer.  By 2015, that had increased to 16.1 percent, a 1.9 percentage point increase. In 2015, two-thirds of Baltimore's 278 neighborhoods had more tree canopy than Franklin Square's 16 percent coverage. 

#### Supporting Code

```{r}

# change
nsa_tree_temp %>%
  select(nsa_name, `avg_canopy_%_07` = `07_lid_mean`, `avg_canopy_15` = `15_lid_mean`, `%_point_change` = lid_change_percent_point) %>%
  filter(nsa_name == "franklin square")

# rank
nsa_tree_temp %>%
  select(nsa_name, `avg_canopy_15` = `15_lid_mean`) %>%
  arrange(nsa_name) %>%
  mutate(rank = dense_rank(desc(`avg_canopy_15`)), pct_nhoods_w_more_tree_canopy = (rank/max(rank))) %>%
  filter(nsa_name == "franklin square")


```
