---
title: "Role of Trees Analysis"
author: "Roxanne Ready | Howard Center for Investigative Journalism"
date: "8/5/2019"
output: html_document
---

## Introduction

This R markdown document describes the methodology and results of a portion of the data analysis we conducted in support of a reporting project examining the effects of tree canopy inequity across the city of Baltimore, especially as it relates to climate change.

## Setup

```{r echo=TRUE, message=FALSE}
#######################
#### Load Packages ####
#######################

library(tidyverse)
require(scales) # For percent labeling on distribution tables
library(DescTools) # For %like% operator
library(here) # For cleaner file path writing

# Turn off scientific notation in RStudio (prevents coersion to character type)
options(scipen = 999)

#########################
#### Store Variables ####
#########################

#### Common save path ####
path_to_data <- "../data/output-data/cleaned/"

###################
#### Load Data ####
###################

blocks_tree_temp_demographics <- read_csv(paste0(path_to_data, "blocks_tree_temp_demographics.csv"))
csa_tree_temp_demographics <- read_csv(paste0(path_to_data, "csa_tree_temp_demographics.csv"))
nsa_tree_temp <- read_csv(paste0(path_to_data, "nsa_tree_temp.csv"))
zcta_tree_temp_demographics <- read_csv(paste0(path_to_data, "zcta_tree_temp_demographics.csv"))
street_trees_nsa_categorized <- read_csv(paste0(path_to_data, "street_trees_nsa_categorized.csv"))

```

## Temperature Analysis

```{r echo=TRUE, message=FALSE}

```

## Tree Canopy Analysis

```{r echo=TRUE, message=FALSE}

```

## Demographics Analysis

```{r echo=TRUE, message=FALSE}

```

## Street Trees Analysis

```{r echo=TRUE, message=FALSE}

```

## Plots and Graphs
```{r echo=TRUE, message=FALSE}

```